Cosmic web alignments with the shape, angular momentum and peculiar velocities of dark matter halos

### 2014-03-13 paper/main.tex
almost there

We study the alignment of dark matter halos with the cosmic web as described by
the tidal and velocity shear fields. We focus on the alignment of their
shape, angular momentum and peculiar velocities. We use a cosmological N-body
simulation that allows to study dark matter halos spanning almost five orders
of magnitude in mass ($10^{9}$-$10^{14}$) \hMsun and use spatial scales
of $(0.5$-$1.0)$\hMpc to define the cosmic web. After varying the
numerical parameters in our experiments we find that the halo shape
and the tidal field web presents the strongest alignment along
filaments and walls with a signal that gets stronger as the halo mass
increases; in the case of the velocity shear only massive halos
$>10^{12}$\hMsun tend to have their shapes perpendicular to
filaments and walls. For the angular momentum we find alignment
signals only for halos more massive than $10^{12}$\hMsun both in the
tidal and velocity shear webs where the preferences are to be
perpendicular to filaments or parallel to walls,
respectively. Finally, the peculiar velocities show a strong alignment
for all halo masses to move along filaments and walls in the tidal
web, while in the velocity shear the same alignment is present but
weaker and only for halos less massive than $10^{12}$\hMsun. Most of
the different signals between the two web description are due to the
fact that the main directions of stronger tidal tension and
velocity shear coincide for halos below $10^{12}$\hMpc but are
perpendicular to each other for halos more massive than this
threshold.  This clearly separates halos into those that live
in sheets and filaments (lower masses) and nodes of the cosmic web. As
a general conclusion, our results clearly show that the two different
algorithms we used to define the cosmic web describe different physical
aspects of non-linear collapse and should be used in a complementary
way according to the problem at hand.
\end{abstract}
\begin{keywords}
methods: N-body simulations, galaxies: haloes, cosmology: theory, dark
matter, large-scale structure of Universe 
\end{keywords}

@@

There is a long observational tradition stuying galactic properties
as a function of their large scale environment
\citep[e.g.][]{Oemler1974,Dressler1980,Pimbblet2002,Gomez2003,Kauffmann2004,Abbas2006,Baldry2006,Park2007,OMill2008,Wilman2010}. In
these situations the environment definition is usually based on
quantities easily accesible to observations such as local number
density or nearest neighbour measurements \citep{Muldrew2012}. With
the advent of large surveys the visibility of the cosmic web is
clear. As a consequence, in order to capture this filamentarity, the
environment definition started to be more complex, involving shear
properties of the galaxy density field or the reconstructuted dark
matter density field \citep[e.g.]{Lee2005,Sousbie2008,MunozCuartas2011,Tempel2014}.

In parallel, numerical simulations can reproduce the web-like
structure of the galaxy distribution in models based on gravitational
instability in a Dark Matter (DM) dominated universe
\citep[e.g.][]{Bond1996,Colberg2005}. Simulations now allow to follow the time
evolution into the non-linear regime of virialized structures which in
turm should host observable galaxies. The discovery in simulations of
gas filaments that feed galaxies is another theoretical hint that
strenghtened the expected connection between galactic properties ant
their place in the cosmic web \citep{Ocvirk2008,Dekel2009}.

With the aid of high resolution simulations it has been established that halos of
a given mass form earlier in denser regions; concentration, spin and
shape also depend on the local density. In the last decade new
algorithms have implemented cosmic web classifications that go beyond
the local density by defining a location to be a peak, sheet, filament
or void depending on the simmetry properties of the local dark matter
distribution
\citep[e.g][]{Hahn2007,Sousbie2008b,Tweb,Gonzalez2010,Noh2011}.

There are analytical frameworks, such as tidal torque theory, that
aim at describing how galaxies gain angular momentum and expect a
correlation of this quantity with the cosmic web. The observational
studies that try to measure such correlations actually try to use as a
proxy the observed shape of galaxies
\citep{Lee2002,Lee2007,Jones2010}. In this respect it is useful to
have a theoretical baseline for the correlations of angular momentum
and shape with the cosmic web.  There is large tradition of alignment
measurements of shape and angular momentum
\citep[e.g.][]{Hahn2007,AragonCalvo2007,Zhang2009,Codis2012,Trowland2013,Libeskind2013}. The
main result of these previous studies is that shape alignment is a
robust measurement regardless of the methods and simulations. On the
other hand, the results for the angular momentum differ in the degree of the alignment.

In the same spirit of describing the place of galaxies within the
cosmic web, there has been an revival of surveys that measure the
cosmic flow patterns in the Local Universe
\citep{Nusser2011,Tully2013}. Assuming the correspondence between the
divergence of the cosmic flow velocity field and the local matter
overdensity (valid in the linear regime) one could construct accurate
maps of the matter density from peculiar velocities
\citep{Courtois2012}. From this perspective it is interesting to look
at the expected alignment of the peculiar velocities.

@@

The structure of this \documentname is the following. In \S
\ref{sec:theory} we present the theoretical antecedents for the
alignment studies we present in this paper. In \S
\ref{sec:nbody} we present the N-body
cosmological simulation and halo catalogs.  Next, we describe in
\S\ref{sec:algorithms} the two web-finding algorithms we use. In
\S\ref{sec:results} we present our main results about the alignment
of shape, spin and peculiar velocities with respect to the cosmic
web. In \S\ref{sec:conclusions} we present our conclusions.

@@

There is abundant literature on the issue of shape and angular momentum
alignment of dark matter halos with respect to the cosmic web. In this
paper we focus our attention on results published during the last
decade that have made use of large N-body dark matter only
cosmological simulations. There are many works that have addressed
this problem using observational data from large surveys such as the
Sloan Digital Sky Survey (SDSS), however we choose to narrow our
discussion to simulation based studies which are prone to comparison.

@@

These alignments are often measured from the distribution of the
$\mu=\vert\cos\theta\vert$ where $\theta$ is the angle between the two axes of
interest. This is often directly measuered as the absolute value of
the dot product between the two unit vectors along the directions
being tested, for instance in the case of angular momentum one would compute
$\mu=\vert\hat{j}\cdot\hat{n}\vert$ In the case of shape alignments the major
axis is the chosen direction to compare against the cosmic web.

@@

In both algorithms we note the three eigenvalues as
$\lambda_1>\lambda_2>\lambda_3$ and their correspoding eivenvectors as
$\hat{e}_1$, $\hat{e}_2$ and $\hat{e}_3$.

@@

\section{Our Numerical Experiments}

@@

However, for simplicity and readibility we report our results in terms
of alignment with respect to filaments and sheets. Given that the
direction along a filament should be defined by the eigenvector
$\hat{e}_3$ corresponding to the smallest eigenvalue $\lambda_3$, a
strong aligmnet along that vector will be reported as an aligment
along filaments. Correspondingly,

@@

\begin{figure*}
\includegraphics[width=0.8\textwidth]{Fig1.pdf}
\caption{Fraction of halos in a mass bin that show a preferential
  alignment with respect to an eigenvector. Each row presents on of
  the three properties studied in this paper: shape (major axis),
  angular momentum and  peculiar velocity. Left (right) column presents the
  results against the Vweb (Tweb). Strong colors refer to $256^3$ grid
  resolutions and lighter colors to a $512^3$ grid. The thick black
  horizontal line at $0.33$ corresponds to the expected fraction for a
  random vector field. 
\label{fig:preferential}}
\end{figure*}

@@

\begin{figure*}
\includegraphics[width=0.75\textwidth]{Fig2.pdf}
\caption{Median of $\cos\theta$ quantifying the shape alignment for
  the Vweb (left) and the Tweb (right) at two different grid
  resolutions. In the upper (lower) panels the angle $\theta$ is
  measured between the halo major axis and the first (third)
  eigenvector.\label{fig:shape_alignment}. }
\end{figure*}

@@

\begin{figure*}
\includegraphics[width=0.75\textwidth]{Fig3.pdf}
\caption{Median of $|\cos\theta|$ quantifying the angular momentum
  alingment for the Vweb (left) and the Tweb (right) for two different
  grid resolutions. In the upper (lower) panels the angle $\theta$ is
  measured between the first (third) eivenvector and the angular
  momentum vector.\label{fig:spin_alignment}} 
\end{figure*}

@@

\begin{figure*}
\includegraphics[width=0.75\textwidth]{Fig4.pdf}
\caption{Median of $\vert\cos\theta\vert$ quantifying the peculiar velocity
  alignment with the Vweb (left) and the Tweb (right) for two different
  grid resolutions. In the upper (lower) panels the angle $\theta$ is
  measured between the first (third) eivenvector and a halo's peculiar
  velocity.
\label{fig:velocity_alignment}} 
\end{figure*}

@@

\begin{figure}
\includegraphics[width=0.40\textwidth]{Fig5.pdf}
\caption{Median of the interweb alignment for the two grid
  resolutions as a function of the dark matter mass corresponding to
  the halos where the measurement was made. The error bars indicate
  the lower and upper quartiles (Figure in color in the web version).
\label{fig:interweb}} 
\end{figure}

@@

We have examined the alignment of shape, angular momentum and peculiar
velocity of dark matter halos with respect to the cosmic web. We use
publicly available data from two algorithms implemented on a large
cosmological N-body simulation to study halo populations spanning five
orders of magnitude in mass. The first algorithm uses the tidal field
and the second the velocity shear, both present local results on
scales of $0.5$\hMpc to $1.0$\hMpc.

We quantify the alignments in two complementary ways. The first one
measures the fraction of halos in a poulation that is preferentially
aligned with either one of the eigenvectors $\vec{e}_1$, $\vec{e}_2$
or $\vec{e}_3$. The second method measures the average value of the
angle between an eigenvector and the quantity of interest. These two
measurements give us a complete picture for the different degrees of
alignment in the cosmic web.

We find that the strongest alignment occurs for the halo shape with
respect to the Tweb. In this case the halos tend to align with the
third eigenvector, $\vec{e}_3$, meaning that they lie along filaments
and walls. This trend is gets stronger as the halo mass
increases and agrees with all the results published so far. Instead,
for the Vweb, there is only an antialignment for halos more massive
than $10^{12}$\hMsun, a result that is presented here for the first
time.

A much weaker alignment signal is present for the angular momentum. In
the Tweb only the most massive halos $>10^{12}$\hMsun are antialigned
with respect to the filaments, while for the Vweb the massive halos
are alignment with sheets. These results broadly agree with the
published literature. Nevertheless, in some publications
\citep{AragonCalvo2007,Hahn2007} there is an alignment signal
reported at lower halo masses $<10^{12}$\hMsun that we do not detect
in our measurements.

A new result from our study is the alignment for the peculiar
velocities. Here we find a relatively strong signal of alignment along
the direction defined by the third eigenvector and perpendicular to
the first. This signal is clear in the Tweb for all masses below
$<10^{13}$\hMsun. This can be interpreted as a flow parallel to walls
and filaments. In the case of the Vweb similar signal, albeit weaker,
is present only for the low mass halos $<10^{12}$\hMsun.

The different behaviour for the alignments of massive halos in the
Tweb and the Vweb was tracked to a corresponding anti-alignment
between the eigenvectors in the two web grids for massive halos
$>10^{12}$\hMsun. For low mass halos the directions defined by the two
webs point in the same direction. This trends can be interpreted as
non-linear effects that appear in the two different physical
descriptions for the cosmic web.

@@

Our study has shown that the alignment properties of dark matter
halos depend on the physical definition of the cosmic web. This shows
that the main aspect of non-linear gravitational collapse might be
revealead easily depending on the choice of the physical context
(i.e. tidal field vs. velocity shear). There is not a better method,
simply differente perspectives. We will explore the potential of these
two different techniques in undertanding the environmental dependence
of galaxy evolution in future work.

@@

@@

In this Appendix we review the results that use a similar exploration
techniques for halo alignments with the cosmic web. Other kind of
alignment statistics based based on modifications of the correlation
\citep{Paz2008,Faltenbacher2009} that go beyond a local computation
and therefore are not reviewed here.

>>>>

We study the alignment of dark matter halos with the cosmic web
charachterized by the tidal and velocity shear fields. We focus on the
alignment of their shape, angular momentum and peculiar velocities. We
use a cosmological N-body simulation that allows to study dark matter
halos spanning almost five orders of magnitude in mass
($10^{9}$-$10^{14}$) \hMsun and use spatial scales of
$(0.5$-$1.0)$\hMpc to define the cosmic web. After varying the
numerical parameters in our experiments we find that the halo shape
presents the strongesalignment along filaments and walls defined by
the tidal field with a signal that gets stronger as the halo mass
increases; in the case of the velocity shear field only massive halos
$>10^{12}$\hMsun tend to have their shapes perpendicular to filaments
and walls. For the angular momentum we find alignment signals only for
halos more massive than $10^{12}$\hMsun both in the tidal and velocity
shear webs where the preferences are to be perpendicular to filaments
or parallel to walls, respectively. Finally, the peculiar velocities
show a strong alignment for all halo masses to move along filaments
and walls in the tidal web, while in the velocity shear the same
alignment is present but weaker and only for halos less massive than
$10^{12}$\hMsun. Most of the different signals between the two web
description are due to the fact that the main directions of stronger
tidal tension and velocity shear coincide for halos below $10^{12}$\hMpc but are
perpendicular to each other for halos more massive than this
threshold.  ur results clearly show that the two different
algorithms we used to define the cosmic web describe different physical
aspects of non-linear collapse and should be used in a complementary
way to understand the effect of the cosmic web on galaxy evolution.
\end{abstract}
\begin{keywords}
methods: N-body simulations, galaxies: haloes, cosmology: theory, dark
matter, large-scale structure of Universe 
\end{keywords}

@@

There is a long observational tradition studying galactic properties
as a function of their large scale environment
\citep[e.g.][]{Oemler1974,Dressler1980,Pimbblet2002,Gomez2003,Kauffmann2004,Abbas2006,Baldry2006,Park2007,OMill2008,Wilman2010}. In
these situations the environment definition is usually based on
quantities easily accesible to observations such as local number
density or nearest neighbour measurements \citep{Muldrew2012}. With
the advent of large surveys the visibility of the cosmic web became
clear. As a consequence, in order to capture its filamentarity, the
environment def nition started to be more complex, involving shear
properties of the galaxy density field or the reconstructuted dark
matter density field
\citep[e.g.]{Lee2005,Sousbie2008,MunozCuartas2011,Tempel2014}.

In parallel to the observational advances, numerical simulations
successfully reproduced the web-like structure of the galaxy
distribution in models based on gravitational instability in a Dark
Matter (DM) dominated universe
\citep[e.g.][]{Bond1996,Colberg2005}. Simulations now allow to follow
the time evolution into the deep non-linear regime of virialized
structures (DM halos) which in turn should host observable
galaxies to study their evolution in the cosmic web. The discovery in
simulations of gas filaments that feed galaxies is another theoretical
hint that strenghtened the expected connection between galactic
properties ant their place in the web \citep{Ocvirk2008,Dekel2009}.

In the last decade new algorithms have implemented cosmic web
classifications that go beyond the local density by defining a
location to be a peak, sheet, filament or void depending on the
simmetry properties of the local dark matter distribution. With the
aid of simulations it has been established that halos of a given mass
form earlier in denser regions; concentration, angular momentum and shape also
depend can be aligned with these web elements. \citep[e.g][]{Hahn2007,Sousbie2008b,Tweb,Gonzalez2010,Noh2011}.

The observational studies that try to measure angular momentum correlations
use the galaxy shape as a proxy  \citep{Lee2002,Lee2007,Jones2010}. In
this respect it is useful to have a theoretical baseline for the
correlations of angular momentum and shape with the cosmic web.  There
is large tradition of alignment measurements of shape and angular momentum
\citep[e.g.][]{Hahn2007,AragonCalvo2007,Zhang2009,Codis2012,Trowland2013,Libeskind2013}. The
main result of these previous studies is that shape alignment is a
robust measurement regardless of the methods and simulations. On the
other hand, the results for the angular momentum differ in the degree
of the alignment.

In the same spirit of describing the place of galaxies within the
cosmic web, there has been an revival of surveys that measure the
cosmic flow patterns in the Local Universe
\citep{Nusser2011,Tully2013}. Assuming the linearity between the
divergence of the cosmic flow velocity field and the local matter
overdensity (valid in the linear regime) one could construct accurate
maps of the matter density from peculiar velocities
\citep{Courtois2012}. From this perspective it is interesting to look
at the expected alignment of the peculiar velocities.

@@

The structure of this \documentname is the following. In \S
\ref{sec:theory} we present the theoretical antecedents for the
alignment studies we present in this paper. In \S
\ref{sec:nbody} we present the N-body
cosmological simulation and halo catalogs.  Next, we describe in
\S\ref{sec:algorithms} the two web-finding algorithms we use and in
\S\ref{sec:experiments} the setup for out numerical experiments. In
\S\ref{sec:results} we present our main results about the alignment
of shape, angular momentum and peculiar velocities with respect to the cosmic
web. In \S\ref{sec:conclusions} we present our conclusions.

@@

@@

In this paper we focus our attention on results published during the last
decade that have made use of large N-body dark matter only
cosmological simulations. There are many works that have addressed
this problem using observational data from large surveys such as the
Sloan Digital Sky Survey (SDSS), however we choose to narrow our
discussion to simulation based studies which are prone to comparison.

Alignments are often measured from the distribution of the
$\mu=\vert\cos\theta\vert$ where $\theta$ is the angle between the two axes of
interest. This is often directly measured as the absolute value of
the dot product between the two unit vectors along the directions
being tested. For instance, in the case of angular momentum one would compute
$\mu=\vert\hat{j}\cdot\hat{n}\vert$. In the case of shape alignments the major
axis is the chosen direction to compare against the cosmic web.

@@

@@

\section{Our Numerical Experiments}
\label{sec:experiments}

@@

However, for simplicity and readibility we report our results in terms
of alignment with respect to filaments and sheets. Given that the
direction along a filament should be defined by the eigenvector
$\hat{e}_3$ corresponding to the smallest eigenvalue $\lambda_3$, a
strong aligmnet along that vector will be reported as an aligment
along filaments. Correspondingly, the first eigenvector $\hat{e}_{1}$
defines the direction perpendicular to walls, a strong alignemnt along
this vector will be reported as an anti-alignment along walls.

@@

\begin{figure*}
\includegraphics[width=0.8\textwidth]{Fig1.pdf}
\caption{Fraction of halos in a mass bin that show a preferential
  alignment with respect to an eigenvector in the cosmic
  web; $\hat{e}_1$ (black) defines the direction perpendicular to a wall and
  $\hat{e}_3$ (yellow) indicates the direction along a filament. Each row
  presents one of the three properties studied in this paper: shape (major axis),
  angular momentum and  peculiar velocity. Left (right) column presents the
  results against the Vweb (Tweb). Strong colors refer to $256^3$ grid
  resolutions and lighter colors to a $512^3$ grid. The thick black
  horizontal line at $0.33$ corresponds to the expected fraction for a
  random vector field. The uncertainty is computed assumed poissonian
  statistics in each mass bin.
\label{fig:preferential}}
\end{figure*}

@@

\begin{figure*}
\includegraphics[width=0.75\textwidth]{Fig2.pdf}
\caption{Median of $|\cos\theta|$ quantifying the shape alignment for
  the Vweb (left) and the Tweb (right) at two different grid
  resolutions as a function of halo mass. In the upper (lower) panels
  the angle $\theta$ is measured between the halo major axis and the first (third)
  eigenvector.\label{fig:shape_alignment}. }
\end{figure*}

@@

\begin{figure*}
\includegraphics[width=0.75\textwidth]{Fig3.pdf}
\caption{Median of $|\cos\theta|$ quantifying the angular momentum
  alingment for the Vweb (left) and the Tweb (right) for two different
  grid resolutions as a function of halo mass. In the upper (lower)
  panels the angle $\theta$ is measured between the first (third)
  eivenvector and the angular momentum
  vector.\label{fig:spin_alignment}}  
\end{figure*}

@@

\begin{figure*}
\includegraphics[width=0.75\textwidth]{Fig4.pdf}
\caption{Median of $\vert\cos\theta\vert$ quantifying the peculiar velocity
  alignment with the Vweb (left) and the Tweb (right) for two different
  grid resolutions as a function of halo mass. In the upper (lower)
  panels the angle $\theta$ is measured between the first (third)
  eivenvector and a halo's peculiar velocity.
\label{fig:velocity_alignment}} 
\end{figure*}

@@

\begin{figure}
\includegraphics[width=0.40\textwidth]{Fig5.pdf}
\caption{Median of the interweb alignment for the two grid
  resolutions as a function of the dark matter mass corresponding to
  the halos where the measurement was made. The error bars indicate
  the lower and upper quartiles.
\label{fig:interweb}} 
\end{figure}

@@

We have examined the alignment of shape, angular momentum and peculiar
velocity of dark matter halos with respect to the cosmic web. We use
publicly available data from two algorithms implemented on a large
cosmological N-body simulation to study halo populations spanning five
orders of magnitude in mass. The first algorithm uses the tidal field (Tweb)
and the second the velocity shear (Vweb), both include results on spatial scales of
$0.5$\hMpc and $1.0$\hMpc.

We quantify the alignments in two complementary ways. The first one
measures the fraction of halos in a poulation that is preferentially
aligned with either one of the eigenvectors $\vec{e}_1$, $\vec{e}_2$
or $\vec{e}_3$ en the local definition of the cosmic web. The second
method measures the average value of the angle between an eigenvector
and the quantity of interest. These two measurements give us a
complete picture for the different degrees of alignment in the web.

We find that the strongest alignment occurs for the halo shape with
respect to the Tweb. In this case the halos tend to align with the
third eigenvector, $\hat{e}_3$, meaning that they lie along filaments
and walls. This trend gets stronger as the halo mass
increases and agrees with all the results published so far. Instead,
for the Vweb, there is only an antialignment for halos more massive
than $10^{12}$\hMsun, a result that is presented here for the first
time.

A much weaker alignment signal is present for the angular momentum. In
the Tweb only the most massive halos $>10^{12}$\hMsun are antialigned
with respect to the filaments, while for the Vweb the massive halos
are aligned with sheets. These results broadly agree with the
published literature. Nevertheless, in some publications
\citep{AragonCalvo2007,Hahn2007} there is an alignment signal
reported at lower halo masses $<10^{12}$\hMsun that we do not detect
in our measurements.

A new result from our study is the alignment for the peculiar
velocities. Here we find a relatively strong signal of alignment along
the direction defined by the third eigenvector, $\hat{e}_3$, and perpendicular to
the first, $\hat{e}_1$. This signal is clear in the Tweb for all masses below
$<10^{13}$\hMsun. This can be interpreted as a flow parallel to walls
and filaments. In the case of the Vweb similar signal, albeit weaker,
is present only for the low mass halos $<10^{12}$\hMsun.

The different behaviour for the alignments of massive halos in the
Tweb and the Vweb is explained by an anti-alignment between the
eigenvectors in the two web grids for massive halos
$>10^{12}$\hMsun. For low mass halos the directions defined by the two
webs point in the same direction. This trends can be interpreted as
non-linear effects that appear in the two different physical
descriptions for the cosmic web.

@@

Our study has shown that the alignment properties of dark matter
halos depend on the physical definition of the cosmic web. This shows
that the main aspect of non-linear gravitational collapse might be
revealead easily depending on the choice of the physical context
(i.e. tidal field vs. velocity shear). There is not a better method,
but differente perspectives. In future work we will explore the
potential of these two complementary techniques in understanding the
environmental dependence of galaxy evolution.

@@

J.E.F-R acknowledges the financial support by Vicerrector\'ia de
Investigaciones at Universidad de los Andes through a FAPA grant.

@@

In this Appendix we review the results that use a similar exploration
techniques for halo alignments with the cosmic web. Other kind of
alignment statistics based based on modifications of the correlation
\citep[e.g.][]{Paz2008,Faltenbacher2009} that go beyond a local computation
and are not reviewed here.
